---
title: "Reproducible Research: Peer-Assessed Project 2"
author: "Christopher Jones"
date: "11/16/2017"
output: 
  html_document:
    echo: true
    code_folding: hide
    keep_md: true
---

## Historical US Storm Data Analysis:
## A survey of historical storm data with respect to population health and economic consequences {.tabset}

### Synopsis:

We examine historical US storm data from NOAA, to assess types of events that are most harmful to human health, and also which types of events have the greatest economic consequences.

Primary findings:  

* Finding 1
* Finding 2
* Finding 3

### Data Loading/Processing

#### Loading:

Loading the data is simple; the code is more lengthy because it is general-purpose, checking for the minimal amount of work to be done to load (need download? need unzip? only need to read the file?).

```{r data_load}
# =======================
# Data loader
# =======================

# download data if it isn't already in place in current directory
filename <- "repdata_data_StormData"
extension.data <- ".csv"
extension.zip <- ".bz2"
zipurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if (!file.exists(paste(filename, extension.data, sep=''))) 
{
  message("Data file not found, checking for data zip file
          ")
  if (!file.exists(paste(c(filename, extension.data, extension.zip), collapse='')))
  {
    message("Data zip file not found. Downloading.")
    
    download <- download.file(zipurl, destfile=paste(c(filename, extension.data, extension.zip), collapse=''))
    foundzip <- TRUE
    
    message("Data zip file downloaded.")
  } else
  {
    foundzip <- TRUE
  }
  
  if (foundzip)
  {
    message("Found data zip file, unzipping.")
    
    library(R.utils)
    bunzip2(paste(c(filename, extension.data, extension.zip), collapse='')
            , paste(c(filename, extension.data), collapse='')
            , remove = FALSE
            , overwrite = TRUE
            )
    unlink(paste(filename, extension.zip, sep=''))
    
    message("Data zip file unzipped.")
  }
  
  if (file.exists(paste(filename, extension.data, sep=''))) 
  {
    message("Data file exists, ready for preprocessing")
  } else
  {
    stop("Data file not found, reason unknown. Halting execution.")
  }
}

# read data only if not already loaded
readdata <- TRUE
if (exists("data.raw"))
{
  if(nrow(data.raw) == 902297) # hardcoding numbers is bad m'kay?
  {
    readdata <- FALSE
  }
}
if (readdata)
{
  message("Reading data file:")
  data.raw <- read.csv(paste(filename, extension.data, sep=''), header=TRUE)
}

str(data.raw)
```

#### Processing:

From the raw data, we perform the following processing steps:

* Subset out the columns needed to analyze storm event consequences on population health and the economy
* Convert dollar value orders of magnitude into actual dollar amounts
* Anything else

```{r data_processing}
data.proc <- data.raw[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]

denominations <- list(list(c('h', 'H'), 2)
                      , list(c('t', 'T'), 3)
                      , list(c('m', 'M'), 6)
                      , list(c('b', 'B'), 9)
                  )

```

### Results: Health

Instructions:

>What is mean total number of steps taken per day?
>
>For this part of the assignment, you can ignore the missing values in the dataset.
>
>Make a histogram of the total number of steps taken each day
>
>Calculate and report the mean and median total number of steps taken per day

The required information is provided by the plot below, using the complete observations dataset, per the instructions.

```{r steps_per_day}
1+1
```


### Results: Economic

Instructions:

>What is the average daily activity pattern?
>
>Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
>
>Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

In addition to answering the questions given, I've provided a smoothing line. The average daily activty pattern is depicted below. On average, the maximum number of steps occurs at interval 835.

```{r daily_activity_pattern}
2+2
```


### References

population: 
	https://seer.cancer.gov/popdata/download.html 
	https://seer.cancer.gov/popdata/yr1969_2015.singleages/us.1969_2015.singleages.adjusted.exe
population dictionary: 
	https://seer.cancer.gov/popdata/popdic.html

fields are parsed by character count, per the following example. see the dictionary link above to interpret the race, etc. codes.

year	state	state	county	seer	race	orig	sex		age		population
				fips	fips	reg
1969	AL		01		001		99		1		9		1		00		00000159


1969AL01001991910100000657
1969AL01001991910200001137
1969AL01001991910300000956

For missing values, we use the predictive mean matching method from the commonly-used mice package (2 data sets, 10 iterations).

Finally, we perform the imputation, and produce a before/after comparative visual:

```{r imputed_values}
3+3
```


fin

